Quantile Regression Analysis for Statin Effects on Body mass Index

### 2021-04-23 manuscript/manuscript2.Rmd
update

 One of the limitations of using polynomial regression is the curvature that can be captured is limited with low order terms. However, including higher-order terms has a negative impact on the model by introducing undesirable "wiggliness" in the regression equation.  Another robust approach to model nonlinear relationships is splines. It is similar to a technique used by draftsmen in plotting curves. The spline is a process of constructing a set of piece-wise continuous
polynomials that are smoothly connected at a set of points in the range of the predictor variable. The connection points are called knots i.e. splines are used to smoothly interpolate between certain points. Let $a,b$ ,and $c$ are the endpoints of the $x$-axis intervals. A smooth cubic spline function is defined by the following equation.
\begin{equation}
f(X)=X\beta=\beta_{0}+\beta_{1}X+\beta_2X^2+\beta_{3}X^{3}+\beta_{4}(X-a)_{+}^3+\beta_{5}(X-b)_{+}^3+\beta_6(X-c)_{+}^3,
\end{equation}

where $$(u)_{+}=  
 \left\{
\begin{array}{ll}
u  & \mbox{if } u > 0 \\
0 & \mbox{if } u \leq 0
\end{array}
\right. $$

@@

 The independent variables in our model are gender, race, age, total cholesterol, cholesterol drug use (yes or no). All types of cholesterol drugs are used including statins. The included races are non-Hispanic white, non-Hispanic black, Hispanic, or other.

@@

```{r resu3, out.width = "80%", fig.align = "center", fig.cap = "Illustration of the quadratic age effect on BMI leveles for four different quantiles of the conditional BMI distribution.  The ", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "Age_poly.png"))) {
    source(here::here("ci_splin.R"))
}
knitr::include_graphics(here::here("images", "Age_poly.png"))
```
Using splines
```{r resu33, out.width = "80%", fig.align = "center", fig.cap = "Age effect modeled using bsplines ", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "Age_splin.png")))
knitr::include_graphics(here::here("images", "Age_splint.png"))
```

>>>>

 One of the limitations of using polynomial regression is the curvature that can be captured is limited with low order terms. However, including higher-order terms has a negative impact on the model by introducing undesirable "wiggliness" in the regression equation.  Another robust approach to model nonlinear relationships is splines. It is similar to a technique used by draftsmen in plotting curves. The spline is a process of constructing a set of piece-wise continuous
<!-- polynomials that are smoothly connected at a set of points in the range of the predictor variable. The connection points are called knots i.e. splines are used to smoothly interpolate between certain points. Let $a,b$ ,and $c$ are the endpoints of the $x-$axis intervals. A transformation for some or all of predictors is needed to capture the nonlinearity in the model. The family of transformation of the predictors that can be fit together to built the model's shape is known as a basis function. the basis functions are $b_1(x),b_2(x), \cdots,\b_k(x),$ and the estimation of $y_i$ is computed as follows: -->

<!-- \begin{equation} -->
<!-- y_i=\beta_0+\beta_1 b_1(x_i)+\beta_2 b_2(x_i)+\cdots+ \beta_k b_k(x_i) -->
<!-- \end{equation} -->
<!-- This concept of a family of transformations that can fit together to capture general shapes is called a basis function. In this case, our objects are functions: $ b_1 (X ), b_2 (X ),\cdots , b_K (X ).$ In a more general way to represent a value of $y=f(X)$ using a peacewise cubic polybomials with a singl knote $c.$  -->

\begin{equation}
y_{i}=  
 \left\{
\begin{array}{ll}
 \beta_{01}+\beta_{11}x_i+\beta_{21}x_i^2+ \beta_{31}x_i^3+	\epsilon_i & \mbox{if } x_i< c \\
 \beta_{02}+\beta_{12}x_i+\beta_{22}x_i^2+ \beta_{23}x_i^3+	\epsilon_i& \mbox{if } x_i \geq c
\end{array}
\right.
\end{equation}
Imposing a continuty condition  and first, and second derivative at $c$ are equal in the two sides, we get a cubic splines. 
<!-- \begin{equation} -->
<!-- f(X)=X\beta=\beta_{0}+\beta_{1}X+\beta_2X^2+\beta_{3}X^{3}+\beta_{4}(X-a)_{+}^3+\beta_{5}(X-b)_{+}^3+\beta_6(X-c)_{+}^3, -->
<!-- \end{equation} -->

<!-- where $$(u)_{+}=   -->
<!--  \left\{ -->
<!-- \begin{array}{ll} -->
<!-- u  & \mbox{if } u > 0 \\ -->
<!-- 0 & \mbox{if } u \leq 0 -->
<!-- \end{array} -->
<!-- \right. $$ -->

@@

 The independent variables in our model are gender, race, age, total cholesterol, cholesterol drug use (yes or no). All types of cholesterol drugs are used including statins. The included races are non-Hispanic white, non-Hispanic black, Hispanic, or other. The following formulas represents the different model that presented in the result.

 \begin{equation}
 \hat{BMI}= bs(Age, df = 5) + Race + Gender + Cholesterol Drug Use
 \end{equation}

 \begin{equation}
 \hat{BMI}= bs(Total chol, df=5)+ Race + Gender + Cholesterol Drug Use
 \end{equation}

 \begin{equation}
 \hat{BMI}= Age+Age^2+ Race + Gender + Cholesterol Drug Use 
 \end{equation}

 \begin{equation}
 \hat{BMI}= Total_chol+ Total_chol^2+ Race + Gender + Cholesterol_Drug_Use
 \end{equation}

@@

```{r resu3, out.width = "80%", fig.align = "center", fig.cap = "Illustration of the quadratic age effect on BMI leveles for four different quantiles of the conditional BMI distribution. ", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "Age_poly.png"))) {
    source(here::here("ci_splin.R"))
}
knitr::include_graphics(here::here("images", "Age_poly.png"))
```
Using splines
```{r resu33, out.width = "80%", fig.align = "center", fig.cap = "Age effect modeled using bsplines ", echo = FALSE}
# Created by file exploratory-data-analysis.R
if (!file.exists(here::here("images", "Age_splin.png"))) {}
knitr::include_graphics(here::here("images", "Age_splin.png"))
```

### 2021-04-23 manuscript/manuscript2.tex
update

Body Mass Index (BMI) plays an important role in predicting heart disease risk(Katzmarzyk et al. 2012). Approximately, 18 million deaths annually per year are caused by cardiovascular diseases(CVD), and similar to this number of nonfatal cardiovascular events (Hay et al. 2017).
In 2011, annual costs for CVD and stroke were \(\$320.1\) billion, which is more than cancer cost. This cost includes \$195.6 billion in direct costs (healthcare costs), and the cost of future productivity loss is \$124.5 (Mozaffarian et al. 2015). Important factors associated with CVD are high body mass index (BMI) and abnormal lipid ratio (Yusuf et al. 2004), (Anderson et al. 1991). Lowering low-density lipoprotein (LDL) by using statin or other drugs reduces the risk of cardiovascular diseases even with a population with no CVD (Yusuf et al. 2016). Ferrieres et al.~(Ferrières et al. 2018) studied the effect of BMI on the choice of lipid-lowering treatment. It is been found that statin intensity increases with the high level of BMI(\(\rho=0.13\)).
\begin{equation}
y_{i}=  
 \left\{
\begin{array}{ll}
 \beta_{01}+\beta_{11}x_i+\beta_{21}x_i^2+ \beta_{31}x_i^3+\eps_i & \mbox{if } x_i< c \\
 \beta_{02} & \mbox{if } x_i \geq c
\end{array}
\right.
\end{equation}
Taking cholesterol drug medication associates with a high risk of elevating BMI level. It is been shown that statin users consume 192 additional calories per day which causes gaining a 6 lb to 11 lb. in a year. Statin users gain 1.3 units in the BMI measure while non-staining users gain 0.4 unit. Moreover, consumption of fat in statin users raised by 14.4\%(Sugiyama et al. 2014).

@@

Least square method expresses conditional mean of y given x as \(\mu(x)=x^T\beta\) and it solves \[ \underset{\beta\in \mathcal{R}^p}{\text{min}}\sum_{i=1}^n(y_i- x_i^T\beta)^2.\] Quantile regression expresses conditional quantile function \(Q_y(\tau|x)=x^T \beta ({\tau)}\) and solve \[ \underset{\beta\in \mathcal{R}^p}{\text{min}}\sum_{i=1}^n\rho_{\tau}(y_i- x_i^T\beta)^2.\]This minimization problem can be reformulated to a linear programming problem \[ \underset{\beta\in \mathcal{R}^p}{\text{min}} \]

@@

One of the limitations of using polynomial regression is the curvature that can be captured is limited with low order terms. However, including higher-order terms has a negative impact on the model by introducing undesirable ``wiggliness'' in the regression equation. Another robust approach to model nonlinear relationships is splines. It is similar to a technique used by draftsmen in plotting curves. The spline is a process of constructing a set of piece-wise continuous
polynomials that are smoothly connected at a set of points in the range of the predictor variable. The connection points are called knots i.e.~splines are used to smoothly interpolate between certain points. Let \(a,b\) ,and \(c\) are the endpoints of the \(x\)-axis intervals. A smooth cubic spline function is defined by the following equation.
\begin{equation}
f(X)=X\beta=\beta_{0}+\beta_{1}X+\beta_2X^2+\beta_{3}X^{3}+\beta_{4}(X-a)_{+}^3+\beta_{5}(X-b)_{+}^3+\beta_6(X-c)_{+}^3,
\end{equation}

where \[(u)_{+}=  
 \left\{
\begin{array}{ll}
u  & \mbox{if } u > 0 \\
0 & \mbox{if } u \leq 0
\end{array}
\right. \]

@@

base\_box\_plot

@@

{\centering \includegraphics[width=0.8\linewidth]{/home/mabdulsa/BMI/images/exploratory2}

\caption{A better figure caption}\label{fig:unnamed-chunk-1}
\end{figure}

So,

@@

\caption{A 90th quantile of BMI ploted with repect to age. The population grouped with repect gender and fasting blood glucose level: prediabetes nad diabetes}\label{fig:unnamed-chunk-2}
\end{figure}

@@

So,

@@

{\centering \includegraphics[width=0.8\linewidth]{/home/mabdulsa/BMI/images/p_age_effect}

@@

\caption{Illustration of the quadratic age effect on BMI leveles for four different quantiles of the conditional BMI distribution.  The }\label{fig:resu3}
\end{figure}

@@

{\centering \includegraphics[width=0.8\linewidth]{/home/mabdulsa/BMI/images/s_age_effect}

@@

Marginal quadratic cholesterol effect

@@

{\centering \includegraphics[width=0.8\linewidth]{/home/mabdulsa/BMI/images/p_TC_effect}

@@

\caption{Illustration of the quadratic cholestrol effect on BMI leveles for four different quantiles of the conditional BMI distribution.  The }\label{fig:resu4}
\end{figure}

Splines modeling

@@

{\centering \includegraphics[width=0.8\linewidth]{/home/mabdulsa/BMI/images/s_TC_effect}

@@

\caption{ Illustration of marginal cholestrol effect modeled as B splines.  The }\label{fig:resu330}
\end{figure}

>>>>

Body Mass Index (BMI) plays an important role in predicting heart disease risk(Katzmarzyk et al. 2012). Approximately, 18 million deaths annually per year are caused by cardiovascular diseases(CVD), and similar to this number of nonfatal cardiovascular events (Hay et al. 2017).
In 2011, annual costs for CVD and stroke were \(\$320.1\) billion, which is more than cancer cost. This cost includes \$195.6 billion in direct costs (healthcare costs), and the cost of future productivity loss is \$124.5 (Mozaffarian et al. 2015). Important factors associated with CVD are high body mass index (BMI) and abnormal lipid ratio (Yusuf et al. 2004), (Anderson et al. 1991). Lowering low-density lipoprotein (LDL) by using statin or other drugs reduces the risk of cardiovascular diseases even with a population with no CVD (Yusuf et al. 2016). Ferrieres et al.~(Ferrières et al. 2018) studied the effect of BMI on the choice of lipid-lowering treatment. It is been found that statin intensity increases with the high level of BMI(\(\rho=0.13\)).

Taking cholesterol drug medication associates with a high risk of elevating BMI level. It is been shown that statin users consume 192 additional calories per day which causes gaining a 6 lb to 11 lb. in a year. Statin users gain 1.3 units in the BMI measure while non-staining users gain 0.4 unit. Moreover, consumption of fat in statin users raised by 14.4\%(Sugiyama et al. 2014).

@@

Least square method expresses conditional mean of y given x as \(\mu(x)=x^T\beta\) and it solves \[ \underset{\beta\in \mathcal{R}^p}{\text{min}}\sum_{i=1}^n(y_i- x_i^T\beta)^2.\] Quantile regression expresses conditional quantile function \(Q_y(\tau|x)=x^T \beta ({\tau)}\) and solve \[ \underset{\beta\in \mathcal{R}^p}{\text{min}}\sum_{i=1}^n\rho_{\tau}(y_i- x_i^T\beta)^2.\]This minimization problem can be reformulated to a linear programming problem

@@

One of the limitations of using polynomial regression is the curvature that can be captured is limited with low order terms. However, including higher-order terms has a negative impact on the model by introducing undesirable ``wiggliness'' in the regression equation. Another robust approach to model nonlinear relationships is splines. It is similar to a technique used by draftsmen in plotting curves. The spline is a process of constructing a set of piece-wise continuous

\begin{equation}
y_{i}=  
 \left\{
\begin{array}{ll}
 \beta_{01}+\beta_{11}x_i+\beta_{21}x_i^2+ \beta_{31}x_i^3+ \epsilon_i & \mbox{if } x_i< c \\
 \beta_{02}+\beta_{12}x_i+\beta_{22}x_i^2+ \beta_{23}x_i^3+ \epsilon_i& \mbox{if } x_i \geq c
\end{array}
\right.
\end{equation}
Imposing a continuty condition and first, and second derivative at \(c\) are equal in the two sides, we get a cubic splines.

@@

@@

@@

\caption{A 90th quantile of BMI ploted with repect to age. The population grouped with repect gender and fasting blood glucose level: prediabetes nad diabetes}\label{fig:unnamed-chunk-1}
\end{figure}

@@

@@

{\centering \includegraphics[width=0.8\linewidth]{/home/mabdulsa/BMI/images/Age_poly}

@@

\caption{Illustration of the quadratic age effect on BMI leveles for four different quantiles of the conditional BMI distribution. }\label{fig:resu3}
\end{figure}

@@

{\centering \includegraphics[width=0.8\linewidth]{/home/mabdulsa/BMI/images/Age_splin}

@@

@@

{\centering \includegraphics[width=0.8\linewidth]{/home/mabdulsa/BMI/images/Total_chol_poly}

@@

\caption{Illustration of a 90th quantile regression of the BMI. The cholesterol term modeled as  quadratic .}\label{fig:resu4}
\end{figure}

@@

{\centering \includegraphics[width=0.8\linewidth]{/home/mabdulsa/BMI/images/TC_splin}

@@

\caption{ Illustration of a 90th quantile regression of the BMI. The cholesterol term modeled  as B splines. }\label{fig:resu330}
\end{figure}
